Limited physiological plasticity and reduced relative variance in physiological rates of ectotherm populations under climate change

### 2024-07-03 ms.qmd
New refs and updated quarto

We show that acclimation of physiological rates and changes in relative variance differ across habitats. Our findings confirm previous results that quantify the different capacity of terrestrial, marine and freshwater ectotherms to acclimate [@Seebacher2015-wi; @Gunderson2015-hz; @morley2019physiological]. Our analysis generally supports that of @Seebacher2015-wi, @Gunderson2015-hz and @morley2019physiological that all show a general inability of terrestrial ectotherms to physiologically acclimate. The change in acclimation $Q_{10}$ we found was similar to @Seebacher2015-wi for freshwater organisms (~17%), but higher in marine ectotherms (decrease of 16% versus ~10% in @Seebacher2015-wi), and opposite in terrestrial ectotherms (increase of ~6% compared to an ~8% decrease in @Seebacher2015-wi), possibly due to our use of new $Q_{10}$ based effect sizes that allow us to control for sampling variance.

More importantly, the change in relative variance in physiological rates across habitats is nearly equivalent (marine) or larger (terrestrial) compared to their capacity to acclimate. Only in freshwater ectotherms did reductions in physiological rates as a result of acclimation coincide with no major change in relative variation. Mechanistically, it is unclear what is contributing to reductions in relative variation in physiological rates at higher temperatures. One possible explanation may be that biochemical reactions are operating more efficiently at higher temperatures, or when acclimated, within a species' normal thermal range. Increased efficiency may result in fewer physiological processes ..... Such effects are expected to be amplified across habitats because of different thermal safety margins, with terrestrial ectotherms being closer to their thermal maxima , and this increased efficiency leads reduced constraints on physiological processes. Our analysis showing that biochemical traits are better capable   These results further highlight the potential vulnerability of terrestrial ectotherms to climate change, while calling into question if acclimation in marine systems will be able to counteract the potentially negative effects of reductions in relative variance. Assuming changes in relative variation in physiological rates are underpinned by genetic variation, and that there is a genetic correlation with fitness, such reductions in physiological variance could limit a populations ability to adapt to future climates [@urban2023and].

While we seen a reduction in relative variation, absolute variation in physiological rates did increase with temperature (@fig-s1) -- albeit to a much lesser extent in terrestrial and marine habitats compared to freshwater -- supporting the general notion that capacity for adaptation may still be limited. As of yet, it is unclear whether a focus on relative or absolute variance should be prioritized or how behavioural strategies of ectotherms might be able to compensate for the general weak acclimation abilities of many ectotherms. For example, terrestrial species have greater opportunity for behavioural microhabitat selection which is expected to decrease the impacts of climate change in thermally heterogeneous environments [@huey2012predicting]. Future theoretical models, that vary both mean and variance, are needed to better understand the broader evolutionary and ecological implications of our findings.

@@

@@

### ***No signature of past climate on capacity for physiological plasticity***

Theoretical evolutionary models predict that plasticity should evolve in populations experiencing greater environmental variability (spatial or temporal), particularly when oscillations are predictable over time to make environmental cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key; @Reed2010]. Higher spatial and temporal heterogeneity in terrestrial habitats [@steele2019comparison] suggests that plasticity should be more likely to evolve in terrestrial environments. However, we did not find evidence of this effect, which is consistent with other meta-analyses of heat tolerance [@barley2021limited; @Gunderson2015-hz] (which we did not consider here). Modelling correct microenvironments for over 200 species across such diverse habitats is challenging. The lack of a relationship between physiological plasticity and environmental variability and predictability could be due to the fact that ectotherms simply seek out microhabitats that make their environment quite stable [@huey2012predicting].  Therefore, it is possible that more fine-scale resolution of the habitats that organisms occupy will elucidate expected patterns. Randomly fluctuating environments have been suggested to select for reduced plasticity, particularly if plastic responses are costly and this has been demonstrated by a number of empirical studies [@Leung2020; @rescan2022experimental; @leung2023phenotypic]. For example, using seed beetles (*Callosobruchus maculatus*), @hallsson2012selection showed that experimentally evolving populations under randomly fluctuating thermal conditions did not show any increase in plasticity, but had reduced plasticity. @Leung2020 also showed reduced plasticity in morphological traits of experimentally evolving algae populations (*Dunaliella salina*) when environments were less predictable [see also @leung2023phenotypic]. While it is unclear how these findings relate directly to acclimation because they integrate multiple forms of plasticity, they do suggest that there are costs to being plastic or that the environmental signals are insufficient to trigger endocrine and epigenetic mechanisms that lead to plasticity when environments are not predictable [@Leung2020; @hendry2016key]. The fact that aquatic ectotherms did show a capacity to acclimate suggests that environmental predictability, which is expected to be higher in aquatic environments, may be more important than environmental variability in driving the evolution of plasticity.

>>>>

We show that acclimation of physiological rates and changes in relative variance differ across habitats. Our findings confirm previous results that quantify the different capacity of terrestrial, marine and freshwater ectotherms to acclimate [@Seebacher2015-wi; @Gunderson2015-hz; @morley2019physiological]. Our analysis generally supports that of @Seebacher2015-wi, @Gunderson2015-hz and @morley2019physiological that all show a general inability of terrestrial ectotherms to physiologically acclimate. The change in acclimation $Q_{10}$ we found was similar to @Seebacher2015-wi for freshwater organisms (~17%), but higher in marine ectotherms (decrease of 16% versus ~10% in @Seebacher2015-wi), and opposite in terrestrial ectotherms (increase of ~6% compared to an ~8% decrease in @Seebacher2015-wi), possibly due to our use of new $Q_{10}$ based effect sizes that allow us to control for sampling variance. Importantly, while marine and freshwater ectotherms were capable of partial acclimation, the effect size was small (amounting to $Q_{10}$ dropping from 1.8 to 1.6), suggesting acclimation will, on average, provide limited capacity to maintain physiological rates in the face of climate change.

Changes in relative variance in physiological rates across habitats was nearly equivalent (marine) or larger (terrestrial) compared to their capacity to acclimate. Only in freshwater ectotherms did acclimation in physiological rates coincide with no change in relative variation. Mechanistically, it is unclear what is contributing to reduction in relative variation in physiological rates at higher temperatures. One possible explanation may be that biochemical reactions are operating more efficiently or are relying on more conformationally stable enzymes at higher temperatures [@tattersall2012coping; @angilletta2009thermal; @somero1995proteins]. Such a mechanism may also help explain the differences in relative variance across habitat types. Terrestrial ectotherms may be relying on more specialised enzymatic variants that operate over a more limited temperature range individually but together allow for organisms to function over a wider thermal range. Needless to say, our results further highlight the potential vulnerability of terrestrial ectotherms to climate change, while calling into question if acclimation in marine systems will be able to counteract the potentially negative effects of reductions in relative variance. Assuming changes in relative variation in physiological rates are underpinned by genetic variation, and that there is a genetic correlation with fitness, such reductions in physiological variance could limit a populations ability to adapt to future climates [@urban2023and].

Importantly, while we seen a reduction in relative variation, absolute variation in physiological rates did increase with temperature (@fig-s1) -- albeit to a much lesser extent in terrestrial and marine habitats compared to freshwater -- supporting the general notion that capacity for adaptation may still be limited. As of yet, it is unclear whether a focus on relative or absolute variance should be prioritized or how behavioural strategies of ectotherms might be able to compensate for the general weak acclimation abilities of many ectotherms. For example, terrestrial species have greater opportunity for behavioural microhabitat selection which is expected to decrease the impacts of climate change in thermally heterogeneous environments [@huey2012predicting]. Future theoretical models, that vary both mean and variance, are needed to better understand the broader evolutionary and ecological implications of our findings.

@@

### ***No signature of past climate on capacity for physiological plasticity***
Theoretical evolutionary models predict that plasticity should evolve in populations experiencing greater environmental variability (spatial or temporal), particularly when oscillations are predictable over time to make environmental cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key; @Reed2010]. Higher spatial and temporal heterogeneity in terrestrial habitats [@steele2019comparison] suggests that plasticity should be more likely to evolve in terrestrial environments. However, we did not find evidence of this effect, which is consistent with other meta-analyses of heat tolerance [@barley2021limited; @Gunderson2015-hz] (which we did not consider here). Modelling correct microenvironments for over 200 species across such diverse habitats is challenging. The lack of a relationship between physiological plasticity and environmental variability and predictability could be due to the fact that ectotherms simply seek out microhabitats that make their environment quite stable [@huey2012predicting].  Therefore, it is possible that more fine-scale resolution of the habitats that organisms occupy will elucidate expected patterns. Randomly fluctuating environments have been suggested to select for reduced plasticity, particularly if plastic responses are costly and this has been demonstrated by a number of empirical studies [@Leung2020; @rescan2022experimental; @leung2023phenotypic]. For example, using seed beetles (*Callosobruchus maculatus*), @hallsson2012selection showed that experimentally evolving populations under randomly fluctuating thermal conditions did not show any increase in plasticity, but had reduced plasticity. @Leung2020 also showed reduced plasticity in morphological traits of experimentally evolving algae populations (*Dunaliella salina*) when environments were less predictable [see also @leung2023phenotypic]. While it is unclear how these findings relate directly to acclimation because they integrate multiple forms of plasticity, they do suggest that there are costs to being plastic or that the environmental signals are insufficient to trigger endocrine and epigenetic mechanisms that lead to plasticity when environments are not predictable [@Leung2020; @hendry2016key]. The fact that aquatic ectotherms did show a capacity to acclimate suggests that environmental predictability, which is expected to be higher in aquatic environments, may be more important than environmental variability in driving the evolution of plasticity.

@@
